Formal Modeling and Verification of a Rate-Monotonic Scheduling Implementation with Real-Time Maude

### 2015-10-22 RMS/submission.tex
finish modifying with Prof. Song's comments

\newcommand{\hide}[1]{\ignorespaces}

@@

Liu and Layland~\cite{DBLP:journals/jacm/LiuL73} gave a sufficient
condition for the schedulability of a set of $n$ tasks scheduled by
RMS: $\displaystyle\Sigma^n_{i=1}C_i/T_i \le n(2^{1/n}-1)$, where
$C_i$ and $T_i$ are the computation time and the period of task
$\tau_i$, respectively. Two main directions on RMS have been explored
since then. One is to relax the assumptions on the original RMS model,
making it applicable on more systems.  For instance,
\cite{DBLP:conf/rtss/LehoczkySS87,DBLP:journals/rts/SpruntSL89,DBLP:conf/rtss/LehoczkyR92,DBLP:journals/tc/StrosniderLS95}
allow aperiodic tasks in the scheduling,
\cite{DBLP:journals/pe/LeungW82,audsley1993deadline} generalize RMS to
be \emph{deadline-monotonic}, \cite{DBLP:journals/tc/ShaRL90} allows
resource sharing among tasks,
\cite{dhall1978real,DBLP:journals/rts/LopezGDG03,DBLP:journals/tpds/LopezDG04,DBLP:journals/tc/BaruahG03}
extend RMS on multiprocessors, and
\cite{DBLP:journals/rts/OhS94,DBLP:journals/rts/GhoshMMS98,DBLP:journals/tpds/BertossiMR99}
enhance fault-tolerance. The other direction is to generate better
schedulablity test conditions for the algorithm and its
extensions~\cite{DBLP:conf/rtss/LehoczkySD89,DBLP:conf/rtss/KuoM91,DBLP:journals/tc/BiniBB03,DBLP:journals/rts/LopezGDG03,DBLP:journals/tc/BaruahG03,gardner1999}. The
RMS algorithm is no doubt of practical importance.

@@

In this paper, we use \emph{Real-Time Maude}, a \emph{rewriting}-based
modeling language and analyzing tool for real-time systems, to model a
realistic implementation of RMS algorithm and then verify some desired
properties on the built model. Based on a realistic implementation,
our model extends the original ideal one proposed
in~\cite{DBLP:journals/jacm/LiuL73}, by allowing computation time of
tasks to be non-constant, and by considering overheads and other
details of the enviroments.

@@

\subsection{Rate-Monotonic Scheduling Algorithm}
\label{ss:rms}
%\subsection{Formulation of the Standard Setting}
A task set consists of \emph{only} $n$ periodic tasks
$\tau_1,\ldots,\tau_n$. Task $\tau_i$ has a period $T_i$ and a
computation time $C_i$. First jobs of all tasks are assumed to be
initiated simultaneously. This means that jobs corresponding to task
$\tau_i$ are initiated at times $kT_i$ with integers $k\ge
0$. Deadlines consist of runability constraints only: the job
initiated at time $kT_i$ has $(k+1)T_i$ as its deadline, which is the
initiation time of the next job. The RMS algorithm chooses the
labeling such that $T_1\le T_2\le \ldots \le T_n$. Consequently,
$\tau_i$ receives priority $i$, assuming smaller numbers have higher
priorities. The following assumptions are made:

(A1) Computation time $C_i$ for each task $\tau_i$ is constant and
does not vary with time.

@@

\subsubsection{Specification}
Real-Time Maude models a system using \emph{modules}. A module
specifies a \emph{real-time rewrite theory} ${\cal R} = (\Sigma, E\cup
A , IR, TR)$, where:
\begin{itemize}
\item $\Sigma$ is an algebraic \emph{signature}, that is, a set of
  declarations of \emph{sorts}, \emph{subsorts} and \emph{function
    symbols}.
\item $(\Sigma, E\cup A)$ is a \emph{membership equational logic
  theory}, with $E$ a set of possibly conditional equations, and $A$ a
  set of equational axioms such as associativiy, commutativity and
  identity.  $(\Sigma, E\cup A)$ models the system's ``static'' states
  as an algebraic data type, and includes a built-in specification of
  a sort \verb|Time|.
\item $IR$ is a set of \emph{labeled conditional rewrite rules}
  specifying the system's local transitions. Each rule has the form
  $[l]~:~t\rightarrow t'\mbox{ \textbf{if} }\bigwedge^n_{j=1}cond_j$,
  where each $cond_j$ is an equality $u_j=v_j$ and $l$ is a
  \emph{label}. Such a rule specifies an \emph{instaneous transition}
  from an instance of $t$ to the corresponding instance of $t'$,
  \emph{provided} the conditions hold.
\item $TR$ is a set of \emph{tick rules} $[l]~:~\{t\}\rightarrow\{t'\}
  \mbox{ \textbf{in time} }r\mbox{ \textbf{if} }cond$ that specify
  \emph{timed transitions} advancing time in the \emph{entire} state
  $t$ by $r$ time units. $IR$ and $TR$ together model the ``dynamic''
  behaviors of the system.
\end{itemize}

Real-Time Maude supports \emph{object-oriented} style specifications.
A class declaration $\texttt{class }C\texttt{ |
}att_1\texttt{:}s_1\texttt{,}\ldots\texttt{,}att_n\texttt{:}s_n$
declares a class $C$ with attributes $att_1$ to $att_n$ of sorts $s_1$
to $s_n$. An \emph{object} of class $C$ in a given state is
represented as a term $\texttt{< } O\texttt{:} C \texttt{ | }
att_1\texttt{:}val_1\texttt{,} \ldots
\texttt{,}att_n\texttt{:}val_n\texttt{ >}$ of sort \verb|Object|,
where $O$, of sort \verb|Oid|, is the object's \emph{identifier}, and
where $val_1$ to $val_n$ are the current values of the attributes
$att_1$ to $att_n$. A \emph{subclass} inherits all the attributes and
rules of its superclasses.

\subsubsection{Formal Analysis}
In this paper, we only consider Real-Time Maude's \emph{linear
  temporal logic model checker}, which analyzes whether \emph{each}
behavior satisfies a temporal logic formula. \emph{State propositions}
are terms of sort \verb|Prop|, and their symantics is defined by
equations $\texttt{ceq } statePattern \texttt{ |= } prop \texttt{ = }
b \texttt{ if } cond$, with $b$ a term of sort \verb|Bool|, stating
that $prop$ evaluates to $b$ in states that are instances of
$statePattern$ when the condition $cond$ holds. These equations
together define $prop$ to hold in all states $t$ where $t \texttt{ |=
} prop$ evaluates to \verb|true|. A temporal logic \emph{formula} is
constructed by state propositions and temporal logic operators such as
\verb|True|, \verb|False|, \verb|~|(negation), \verb|/\|, \verb|\/|,
\verb|->|(implication), \verb|[]|(``always''),
\verb|<>|(``eventually'') and \verb|U|(``until''). Real-Time Maude
provides both \emph{unbounded} and \emph{time-bounded} LTL model
checking. The time-bounded model checking command
\begin{alltt}
  (mc \(t\) |=t \(formula\) in time <= \(timeLimit\) .)
\end{alltt}
checks whether the temporal logic formula holds in all behaviors up to
duration $timeLimit$ starting from the initial state $t$.

\section{The Implementation of RMS}
\label{s:imp}
The algorithm of the RMS is straightforward, and it is easy to
implement it in pratical systems. However, the way to implement it is
not unique. Our target implementation to analyze is part of a
realistic avionic operating system from our industrial
partner. Interrupts triggered by the clock would come every $T$, which
we call \emph{interrupt cycle}. When an interrupt request comes, if
the system is interruptable, i.e. the interrupt mask is cleared, the
function $schedule()$ will be called; otherwise the handler function
$schedule()$ will be pending until the interrupt mask becomes cleared.
The pseudocode of $schedule()$ is shown as Algorithm~\ref{a:schedule},
where $taskList$ is the list of periodic tasks, which are needed to be
scheduled, and assumed ordered descendingly by priorities, and where
both variables $taskList$ and $timer$ are global. In this
implementation, there is only one kind of interrupt, the period $T_i$
of each task is a multiple of $T$, and the tasks are independent,
meeting assumption~(A2).

\begin{algorithm}
  \caption{The C-Like Pseudocode of $schedule()$}
  \label{a:schedule}
  \begin{algorithmic}[1]
\Function{$schedule$}{$ $}{}
  \State \Call{$int\_o\!f\!\!f$}{$ $}; \Comment{to disable interrupts}
  \State \Call{$updateStatus$}{$taskList$}; \label{l:updatestatus}
  \State $timer = timer + 1$; \label{l:timer} \label{l:inc}
  \State $p = taskList$;
  \While{$p$} \label{l:startrun1st}
    \If{$p\rightarrow status == \textit{INTERRUPT}$}
      \State \Return;
    \ElsIf{$p\rightarrow status == \textit{READY}$}      
      \State $p\rightarrow status = \textit{RUNNING}$;
      \State \Call{$int\_on$}{$ $}; \Comment{to enable interrupts} \label{l:endrun1st}
      \State $p\rightarrow function()$; \Comment{to run the task} \label{l:function}
      \State \Call{$int\_o\!f\!\!f$}{$ $};
      \State $p\rightarrow status = \textit{DORMANT}$;
    \EndIf
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
\Function{$updateStatus$}{$p$}
  \While{$p$}
    \If{$p\rightarrow status == \textit{RUNNING}$} \label{l:startupdate}
      \State $p\rightarrow status = \textit{INTERRUPT}$;
    \EndIf
    \If{$timer~\%~(p\rightarrow period) == 0$} \Comment{the task should be initiated}
      \If{$p\rightarrow status == \textit{DORMANT}$} 
        \State $p\rightarrow status = \textit{READY}$;
      \Else \Comment{the status is \textit{READY} or \textit{INTERRUPT}}
	\State \Call{$reportTaskError$}{$p$}; \Comment{the task misses its deadline}
      \EndIf
    \EndIf \label{l:endupdate}
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
  \end{algorithmic}
\end{algorithm}

The function $updateStatus()$ updates status of all tasks in the list:
if the task is running when the interrupt is handled, it becomes
interrupted\footnote{Note that the status \textit{INTERRUPT} indicates
  the task is interrupted for the moment, or was interrupted before
  but is not finished running yet.}; the task, which is at its
initiated time, would be set to be ready if its previous corresponding
job is finished, or produce an error otherwise. Notice that
$schedule()$ is called only when the periodic interrupt is handled. It
can be also interrupted while running some task (Line~\ref{l:function}
in Algorithm~\ref{a:schedule}), such that the running of $schedule()$
can be nested.

\section{Formalizing the RMS Implementation}
\label{s:formalism}
Based on the code, we model the implementation under the following
assumptions:

(A1') Computation time $C_i$ for each task $\tau_i$ is not constant,
varying randomly between minimum $C^{min}_i$ and maximum $C^{max}_i$.

@@

(A3') The times for scheduling and task switching should be
considered, while other overhead is ignored.

These assumptions make our model different with the standard
formulation. For instance, due to the interrupt masking and
consideration of scheduling overhead, jobs corresponding to task
$\tau_i$ could not be initiated at exactly $kT_i$, but a bit later.
Note that (A2) says jobs are ready to run at their initiation times,
however, they cannot run at their initiation times, because scheduling
takes time.

\subsection{Some Basic Data Types}
In our model, tasks are identified with their indexes of sort
\verb|Nat| in the $taskList$. We define a sort \verb|MaybeNat|
wrapping \verb|Nat|s to refer to some task, with \emph{constructor}
\verb|some| followed by a \verb|Nat| $n$ indicating the task with
index $n$, and \verb|none| for no task:
\begin{verbatim}
  op none : -> MaybeNat [ctor] .
  op some_ : Nat -> MaybeNat [ctor] .
\end{verbatim}

@@

\subsection{Modeling the System States}
Abstracting from some details, the running system could be considered
as consisting of several parts: the tasks which are scheduled, the
scheduler itself, the hardware including registers and stacks, and the
interrupt source. The scheduler, i.e. the function $schedule()$, has
only the variable $timer$ to be interesting, thus it can be described
by a single variable. We will present the modeling of the other parts
one by one.

\subsubsection{Tasks}
In this paper, since we are trying to analyze the implementation of
the scheduler, each task will be abstracted from its functionality and
treated as a \verb|Counter|. As mentioned above, times for scheduling
and task switching will be also considered in the model. They are just
seen as two system tasks. Every task is modeled as an object instance
of some subclass of the base class \verb|Task|:
\begin{verbatim}
  class Task | cnt : Counter .
  op error : -> Object [ctor] .
\end{verbatim}
The attribute \verb|cnt| denotes the computation time of the task,
while \verb|error| is an object indicating some task which misses its
deadline.

A periodic task, which is needed to be scheduled, is an object
instance of the subclass \verb|PTask| with additional attributes
\verb|priority|, \verb|period| and \verb|status|:
\begin{verbatim}
  class PTask | priority : Nat, period : NzNat, status : Status .
  subclass PTask < Task .
\end{verbatim}
where \verb|Status| is a sort with four constant constructors, just as
in the implementation:
\begin{verbatim}
  ops RUNNING INTERRUPT READY DORMANT : -> Status [ctor] .
\end{verbatim}
The list of periodic tasks, the variable $taskList$ in the
implementation, is modeled as an instance of the sort \verb|TaskList|,
which is a list of \verb|PTask|s\footnote{Following the Maude
  convention, variables would be written in capital letters.}:
\begin{verbatim}
  op null : -> TaskList [ctor] .
  op _::_ : Object TaskList ~> TaskList [ctor] .
  mb (< O:Oid : PTask |> :: L:TaskList) : TaskList .
  mb (error :: L:TaskList) : TaskList .
\end{verbatim}
As mentioned already, periodic tasks would be identified by their
indexes in the list.

On the other hand, a system task is an object instance of the subclass
\verb|SysTask| of \verb|Task| with no extra attributes:
\begin{verbatim}
  class SysTask .            subclass SysTask < Task .
\end{verbatim}
Different from periodic tasks, system tasks would be organized in a
multiset of sort \verb|SysTasks|, and identified by their \verb|Oid|s.
We abstract here from the detailed definitions thanks to the
similarity.

\subsubsection{Hardware} Our model considers two parts of hardware relating
to the interrupt handling inside the system, the registers and the
stack.

The set of registers is modeled as an object instance of the class
\verb|Regs|, with attributes \verb|pc| denoting the program counter,
\verb|mask| for the interrupt masking flag and \verb|ir| for the
interrupt request flag.
\begin{verbatim}
  class Regs | pc : TaskID, mask : Bool, ir : Bool .
\end{verbatim}
where the sort \verb|TaskID| is a supersort of \verb|MaybeNat| and
\verb|Oid|, referring to some task.
\begin{verbatim}
  subsorts MaybeNat Oid < TaskID .
\end{verbatim}

Some operations, such as \verb|getPc|, \verb|setMask| and so on, are
also defined on the class \verb|Regs|.

@@

\subsubsection{Interrupt Source}
The interrupt source is modeled in our model as an object of class \verb|IntSrc|,
with attribute \verb|cycle| denoting the interrupt cycle:
\begin{verbatim}
  class IntSrc | val : Time, cycle : Time .
\end{verbatim}

\subsubsection{System}
The system in the model is just a composition of the parts introduced
above, with a sort \verb|System|\footnote{Some variable declarations
  are not shown for simplicity.}:
\begin{verbatim}
  op _____ : 
       TaskList Timer SysTasks Hardware Object ~> System [ctor] .
  mb (L T STS HW < O:Oid : IntSrc |>) : System .
\end{verbatim}

\subsection{Interrupt Requesting}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero| while time
advances. This is modeled by the following instaneous conditional rule
applied on \verb|System|:
\begin{verbatim}
  crl [interrupt-request] :
    (L T STS HW ISRC) => (L T STS (HW).intReq reset(ISRC))
    if (ISRC).timeout .
\end{verbatim}
where the operator \verb|_.timeout| examines whether the attribute
\verb|val| equals to \verb|zero|, and \verb|_.intReq| sets the
\verb|ir| flag indicating there is an interrupt request to be handled:
\begin{verbatim}
  op _.intReq : Hardware -> Hardware .
  eq [ REGS ; S ].intReq = [ (REGS).setIr ; S ] .
\end{verbatim}
Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

>>>>

\newcommand{\hide}[1]{\ignorespaces}
\newcommand{\jx}[1]{{\bf Jiaxiang: }#1{ \bf End}}

@@

Liu and Layland~\cite{DBLP:journals/jacm/LiuL73} gave a sufficient
condition for the schedulability of a set of $n$ tasks scheduled by
RMS: $\displaystyle\Sigma^n_{i=1}C_i/T_i \le n(2^{1/n}-1)$, where
$C_i$ and $T_i$ are the \emph{computation (time) requirement} and the
period of task $\tau_i$, respectively. Two main directions on RMS have
been explored since then. One is to relax the assumptions on the
original RMS model, making it applicable on more systems.  For
instance,
\cite{DBLP:conf/rtss/LehoczkySS87,DBLP:journals/rts/SpruntSL89,DBLP:conf/rtss/LehoczkyR92,DBLP:journals/tc/StrosniderLS95}
allow aperiodic tasks in the scheduling,
\cite{DBLP:journals/pe/LeungW82,audsley1993deadline} generalize RMS to
be \emph{deadline-monotonic}, \cite{DBLP:journals/tc/ShaRL90} allows
resource sharing among tasks,
\cite{dhall1978real,DBLP:journals/rts/LopezGDG03,DBLP:journals/tpds/LopezDG04,DBLP:journals/tc/BaruahG03}
extend RMS on multiprocessors, and
\cite{DBLP:journals/rts/OhS94,DBLP:journals/rts/GhoshMMS98,DBLP:journals/tpds/BertossiMR99}
enhance fault-tolerance. The other direction is to generate better
schedulablity test conditions for the algorithm and its
extensions~\cite{DBLP:conf/rtss/LehoczkySD89,DBLP:conf/rtss/KuoM91,DBLP:journals/tc/BiniBB03,DBLP:journals/rts/LopezGDG03,DBLP:journals/tc/BaruahG03,gardner1999}. The
RMS algorithm is no doubt of practical importance.

@@

In this paper, we use \emph{Real-Time Maude}, a \emph{rewriting}-based
modeling language and analyzing tool for real-time systems, to model a
realistic implementation of RMS algorithm and then verify some desired
properties on the built model. Based on a realistic implementation,
our model extends the original ideal one proposed
in~\cite{DBLP:journals/jacm/LiuL73}, by allowing computation
requirements of tasks to be non-constant, and by considering overhead
and other details of the enviroments.

@@

\subsection{Rate-Monotonic Scheduling Algorithm}
\label{ss:rms}
%\subsection{Formulation of the Standard Setting}
A task set consists of \emph{only} $n$ periodic tasks
$\tau_1,\ldots,\tau_n$. Task $\tau_i$ has a period $T_i$ and a
computation requirement $C_i$. First jobs of all tasks are assumed to
be initiated simultaneously. This means that jobs corresponding to
task $\tau_i$ are initiated at times $kT_i$ with integers $k\ge
0$. Deadlines consist of runability constraints only: the job
initiated at time $kT_i$ has $(k+1)T_i$ as its deadline, which is the
initiation time of the next job. The RMS algorithm chooses the
labeling such that $T_1\le T_2\le \ldots \le T_n$. Consequently,
$\tau_i$ receives priority $i$, assuming smaller numbers have higher
priorities. The following assumptions are made:

(A1) Computation requirement $C_i$ for each task $\tau_i$ is constant
and does not vary with time.

@@

\subsubsection{Specification}
Real-Time Maude models a system using \emph{modules}. A module
specifies a \emph{real-time rewrite theory} ${\cal R} = (\Sigma, E\cup
A , IR, TR)$, where:
\begin{itemize}
\item $\Sigma$ is an algebraic \emph{signature}, that is, a set of
  declarations of \emph{sorts}, \emph{subsorts} and \emph{function
    symbols}. The function symbols are allowed to be mixfix.
\item $(\Sigma, E\cup A)$ is a \emph{membership equational logic
  theory}, with $E$ a set of conditional equations on $\Sigma$, and
  $A$ a set of equational axioms such as associativiy, commutativity
  and identity.  $(\Sigma, E\cup A)$ models the system's ``static''
  states as an algebraic data type, and includes a built-in
  specification of a sort \verb|Time|.
\item $IR$ is a set of \emph{labeled conditional rewrite rules}
  specifying the system's local transitions. Each rule has the form
  $[l]~:~t\rightarrow t'\mbox{ \textbf{if} }\bigwedge^n_{j=1}cond_j$,
  where each $cond_j$ is an equality $u_j=v_j$ and $l$ is a
  \emph{label}. Such a rule specifies an \emph{instaneous transition},
  without consuming time, from an instance of $t$ to the corresponding
  instance of $t'$, \emph{provided} the conditions hold.
\item $TR$ is a set of (\emph{labeled}) \emph{tick rules}
  $[l]~:~\{t\}\rightarrow\{t'\} \mbox{ \textbf{in time} }r\mbox{
  \textbf{if} }cond$ that specify \emph{timed transitions} advancing
  time in the \emph{entire} state $t$ by $r$ time units. $IR$ and $TR$
  together model the ``dynamic'' behaviors of the system.
\end{itemize}

Real-Time Maude supports specifications in \emph{object-oriented}
style.  A class declaration $\texttt{class }C\texttt{ |
}att_1\texttt{:}s_1\texttt{,}\ldots\texttt{,}att_n\texttt{:}s_n$
defines a class $C$ with attributes $att_1$ to $att_n$ of sorts $s_1$
to $s_n$, respectively. An \emph{object} of class $C$ is represented
as a term $\texttt{< } O\texttt{:} C \texttt{ | }
att_1\texttt{:}val_1\texttt{,} \ldots
\texttt{,}att_n\texttt{:}val_n\texttt{ >}$ of sort \verb|Object|,
where $O$, of sort \verb|Oid|, is the object's \emph{identifier}, and
$val_1$ to $val_n$ are the values of the attributes $att_1$ to
$att_n$. A \emph{subclass} inherits all the attributes and rules of
its superclasses.

\subsubsection{Formal Analysis}
Real-Time Maude provides many useful commands and tools to analyze a
model. For example, \verb|rewrite| allows to execute the model,
symbolically; given an initial state, \verb|search| is used to search
reachable states satisfying the desired properties; the Maude's
\emph{Inductive Theorem Prover} (ITP) can be applied to interactively
prove properties written in \emph{membership equational logic}.

In this paper, we only consider Real-Time Maude's \emph{linear
  temporal logic model checker}, which analyzes whether \emph{each}
behavior satisfies a temporal logic formula. \emph{State propositions}
are terms of sort \verb|Prop|. Their symantics is defined by equations
of the form $\texttt{ceq } statePattern \texttt{ |= } prop \texttt{ =
} b \texttt{ if } cond$, with $b$ a term of sort \verb|Bool|, stating
that $prop$ evaluates to $b$ in states which are instances of
$statePattern$ provided the condition $cond$ holds. These equations
together define $prop$ to hold in all states $t$ that make $t \texttt{
  |= } prop$ evaluate to \verb|true|. A temporal logic \emph{formula}
is constructed by state propositions and temporal logic operators such
as \verb|True|, \verb|False|, \verb|~|(negation), \verb|/\|,
\verb|\/|, \verb|->|(implication), \verb|[]|(``always''),
\verb|<>|(``eventually'') and \verb|U|(``until''). Real-Time Maude
supports both \emph{timed} and \emph{untimed} LTL model checking. The
untimed model checking command
\begin{alltt}
  (mc \(t\) |=u \(formula\) .)
\end{alltt}
checks whether the temporal logic $formula$ holds in all behaviors
starting from the initial state $t$, \emph{with no time limit}.

\section{The Implementation of RMS}
\label{s:imp}
The implementation is in a realistic avionic operating system.
Interrupts triggered by the clock would come every $T$, which we call
\emph{interrupt cycle}. When an interrupt request occurs, if the
system is interruptable, i.e. the interrupt mask is cleared, the
handler function $schedule()$ will be invoked; otherwise $schedule()$
will be pending until the interrupt mask becomes cleared.  The
implementation is shown as $schedule()$ in Algorithm~\ref{a:schedule},
where $taskList$ is the list of periodic tasks to be scheduled. We assume that 
the list is ordered descendingly by priorities, and both variables
$taskList$ and $timer$ are global. In this implementation, there is
only one kind of interrupt, the period $T_i$ of each task is a
multiple of $T$, and the tasks are independent, meeting
assumption~(A2).

\begin{algorithm}
  \caption{The C-Like Pseudocode of $schedule()$}
  \label{a:schedule}
  \begin{algorithmic}[1]
\Function{$schedule$}{$ $}{}
  \State \Call{$int\_o\!f\!\!f$}{$ $}; \Comment{to disable interrupts} \label{l:1stline}
  \State \Call{$updateStatus$}{$taskList$}; \label{l:updatestatus}
  \State $timer = timer + 1$; \label{l:timer} \label{l:inc}
  \State $p = taskList$;
  \While{$p$} \label{l:startrun1st}
    \If{$p\rightarrow status == \textit{INTERRUPT}$}
      \State \Return;
    \ElsIf{$p\rightarrow status == \textit{READY}$}      
      \State $p\rightarrow status = \textit{RUNNING}$;
      \State \Call{$int\_on$}{$ $}; \Comment{to enable interrupts} \label{l:endrun1st}
      \State $p\rightarrow function()$; \Comment{to execute the task} \label{l:function}
      \State \Call{$int\_o\!f\!\!f$}{$ $};
      \State $p\rightarrow status = \textit{DORMANT}$;
    \EndIf
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
\Function{$updateStatus$}{$p$}
  \While{$p$}
    \If{$p\rightarrow status == \textit{RUNNING}$} \label{l:startupdate}
      \State $p\rightarrow status = \textit{INTERRUPT}$;
    \EndIf
    \If{$timer~\%~(p\rightarrow period) == 0$} \Comment{the task should be initiated}
      \If{$p\rightarrow status == \textit{DORMANT}$} \Comment{the previous job finishes} 
        \State $p\rightarrow status = \textit{READY}$;
      \Else \Comment{the status is \textit{READY} or \textit{INTERRUPT}}
	\State \Call{$reportTaskError$}{$p$}; \Comment{the task misses its deadline}
      \EndIf
    \EndIf \label{l:endupdate}
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
  \end{algorithmic}
\end{algorithm}

In Algorithm~\ref{a:schedule}, the handler function $scheduler()$
first updates status of all tasks in $taskList$ via function
$updateStatus()$. Then it scans the list to execute the ready tasks
one by one, or to do a return when encountering an interrupted task.
The function $updateStatus()$ updates each task in two stages:
firstly, if the task is running, it becomes interrupted\footnote{Note
  that the status \textit{INTERRUPT} indicates the task is interrupted
  for the moment, or was interrupted before but its execution is not
  finished yet.}; secondly for the task at its initiation time, if its
previous job is finished, it would be set to be ready, otherwise it
produces an error. Notice that $schedule()$ is invoked only when the
periodic interrupt is handled. Its running cannot be interrupted when
it is updating status of tasks or finding the next task to be
executed, however, it can be interrupted while executing some task
(Line~\ref{l:function}). This allows the running of $schedule()$ to be
nested.

In the rest of this paper, we use \emph{scheduling} to refer to the
stage, from the moment when a pending interrupt request is detected,
to the moment when the first should-be-run periodic task starts
executing, i.e. Line~\ref{l:function} in
Algorithm~\ref{a:schedule}. Therefore, \emph{scheduling time} mainly
consists of three parts: (i) the time for switching context from the
running task, possibly none, to $schedule()$ when handling interrupt
requests, (ii) the time spent by $schedule()$ finding and setting the
\emph{first} should-be-run periodic task
(Lines~\ref{l:1stline}-\ref{l:endrun1st} in
Algorithm~\ref{a:schedule}), and (iii) the time for switching context
from $schedule()$ to that first should-be-run task. \emph{Switching}
refers to the stage, from the moment when a periodic task completes
its execution, to the moment when the next should-be-run periodic task
starts executing. \emph{Switching time} thus also consists of three
parts: (i) the time for switching context from the finished task back
to $schedule()$, (ii) the time spent by $schedule()$ finding and
setting the \emph{next} should-be-run periodic task, and (iii) the
time for switching context from $schedule()$ to that next
should-be-run task.

\section{Formal Model of the Implementation}
\label{s:formalism}
We model the implementation under the following assumptions:

(A1') Computation requirement $C_i$ for each task $\tau_i$ is not
constant. It varies randomly between minimum $C^{min}_i$ and maximum
$C^{max}_i$, i.e. within an interval $[C^{min}_i, C^{max}_i]$.

@@

(A3') Scheduling time and switching time are considered, while other
overhead is ignored.

These assumptions make our model different from the standard one. For
instance, if an interrupt request occurs while switching is performed,
jobs corresponding to task $\tau_i$ could not be initiated at exactly
$kT_i$. They should wait until switching is finished and the interrupt
mask is cleared. Note that (A2) says that jobs are ready to run at
their initiation times, however, no one can start running at exactly
its initiation time, because scheduling takes time.

\jx{TODO. A figure is needed here to demostrate the scheduling under
  the three assumptions above.}

In this section, we introduce first how we model the states--the
static aspect--of the system using algebraic data types, then how we
specify the essential behaviors--the dynamic aspect--using rewrite
rules.

\subsection{Basic Data Types}
In our model, tasks are identified with their indexes of sort
\verb|Nat| in the $taskList$. We define a sort \verb|MaybeNat|
wrapping \verb|Nat|s to refer to some task, with \emph{constructor}
\verb|some| followed by a \verb|Nat| $n$ indicating the task with
index $n$, and \verb|none| for no task:
\begin{verbatim}
  op none : -> MaybeNat [ctor] .
  op some_ : Nat -> MaybeNat [ctor] .
\end{verbatim}

@@

\subsection{Modeling the System States}
The running system can be considered as consisting of several parts:
the tasks which are scheduled, the scheduler itself, the hardware
including registers and stacks, and the interrupt source. The
scheduler, i.e. the function $schedule()$, can be described by a
single variable $timer$. We present the models of the other parts one
by one.

\subsubsection{Tasks}
Each task is abstracted from its functionality as a \verb|Counter|.
Overhead for scheduling and switching is considered in the model. They
are treated as two system tasks. Every task is modeled as an object
instance of some subclass of the base class \verb|Task|:
\begin{verbatim}
  class Task | cnt : Counter .
  op error : -> Object [ctor] .
\end{verbatim}
The attribute \verb|cnt| denotes the computation time of the task,
while \verb|error| is an object indicating some task which misses its
deadline.

A periodic task, which is needed to be scheduled, is an object
instance of the subclass \verb|PTask| with additional attributes
\verb|priority|, \verb|period| and \verb|status|:
\begin{verbatim}
  class PTask | priority : Nat, period : NzNat, status : Status .
  subclass PTask < Task .
\end{verbatim}
where \verb|Status| is a sort with four constant constructors, just as
in the implementation:
\begin{verbatim}
  ops RUNNING INTERRUPT READY DORMANT : -> Status [ctor] .
\end{verbatim}
The list of periodic tasks, the variable $taskList$ in the
implementation, is modeled as an instance of the sort \verb|TaskList|,
which is a list of \verb|PTask|s\footnote{Following the Maude
  convention, variables would be written in capital letters.}:
\begin{verbatim}
  op null : -> TaskList [ctor] .
  op _::_ : Object TaskList ~> TaskList [ctor] .
  mb (< O:Oid : PTask |> :: L:TaskList) : TaskList .
  mb (error :: L:TaskList) : TaskList .
\end{verbatim}
Periodic tasks are identified by their indexes in the list.

On the other hand, a system task is an object instance of the subclass
\verb|SysTask| of \verb|Task| with no extra attributes:
\begin{verbatim}
  class SysTask .            subclass SysTask < Task .
\end{verbatim}
Different from periodic tasks, system tasks are organized in a
multiset of sort \verb|SysTasks|, and identified by their \verb|Oid|s.
%We abstract here from the detailed definitions thanks to the
%similarity.

\subsubsection{Hardware}
Our model considers two parts of hardware related to the interrupt
handling inside the system: the registers and the stack.

The set of registers is modeled as an object instance of the class
\verb|Regs|, with attributes \verb|pc| denoting the program counter,
\verb|mask| for the interrupt masking flag and \verb|ir| for the
interrupt request flag, respectively.
\begin{verbatim}
  class Regs | pc : TaskID, mask : Bool, ir : Bool .
\end{verbatim}
where the sort \verb|TaskID| is a supersort of \verb|MaybeNat| and
\verb|Oid|, referring to some task.
\begin{verbatim}
  subsorts MaybeNat Oid < TaskID .
\end{verbatim}
Some operations, such as \verb|getPc| and \verb|setMask|, are defined
on the class \verb|Regs|.

@@

\subsubsection{Interrupt Source}
The interrupt source is modeled as an object of class \verb|IntSrc|,
with attributes \verb|cycle| denoting the interrupt cycle $T$, and
\verb|val| for the value which will decrease from $T$ to $0$ while
time advances:
\begin{verbatim}
  class IntSrc | val : Time, cycle : Time .
\end{verbatim}

\subsubsection{System}
The entire system in the model is a composition of the parts
introduced above, with a sort \verb|System|\footnote{Some variable
  declarations are not shown for simplicity.}:
\begin{verbatim}
  op _____ : 
       TaskList Timer SysTasks Hardware Object ~> System [ctor] .
  mb (L T STS HW < O:Oid : IntSrc |>) : System .
\end{verbatim}

\subsection{Interrupt Requests}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero|. The
requesting is an instaneous action, and thus is modeled by the
following instaneous conditional rule applied on \verb|System|:
\begin{verbatim}
  crl [interrupt-request] :
    (L T STS HW ISRC) => (L T STS (HW).intReq reset(ISRC))
    if (ISRC).timeout .
\end{verbatim}
where the operator \verb|_.timeout| examines whether the attribute
\verb|val| equals \verb|zero|, and \verb|_.intReq| sets the \verb|ir|
flag indicating there is an interrupt request to be handled:
\begin{verbatim}
  op _.intReq : Hardware -> Hardware .
  eq [ REGS ; S ].intReq = [ (REGS).setIr ; S ] .
\end{verbatim}
Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

\subsection{Task Initiation}
Periodic tasks are initiated sequentially by the function
$updateStatus()$ in Algorithm~\ref{a:schedule}, which is modeled by
the recursive operation as below:
\begin{verbatim}
  op updateStatus_with_ : TaskList Timer -> TaskList . 
  eq updateStatus null with TIMER = null .
  eq updateStatus (TASK :: L) with TIMER
       = (update TASK with TIMER) :: (updateStatus L with TIMER) .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$,
and \verb|update_with_| updating the status of individual task
(Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Algorithm~\ref{a:schedule})
\begin{verbatim}
  op update_with_ : Object Timer ~> Object .
  ceq update < O : PTask | period : T, status : ST > with TIMER
        = if ST == DORMANT then < O : PTask | status : READY >
          else error fi
      if TIMER rem T == 0 .
  eq update < O : PTask | status : ST > with TIMER
       = if ST == RUNNING then < O : PTask | status : INTERRUPT >
         else < O : PTask |> fi [owise] .
\end{verbatim}
In the case where \verb|TIMER|($timer$) can be divided by the task's
period, i.e. the task should be initiated, if the status is
\verb|DORMANT|, it should be ready to run; otherwise it means the
previous job of the task is not finished, thus it misses its deadline,
producing an \verb|error|. In the other case where the tasks should
not be initiated, the status of the task changes only if it is
\verb|RUNNING|. We can see that \verb|updateStatus_with_| behaves the
same as $updateStatus()$.
